### 2025-09-09 getting-started.qmd
20250909 - tweaks

The book uses the software `R` [@R-base] for statistical analyses (<http://www.r-project.org>).
R is a free software environment; you can download it at no charge here: <https://cran.r-project.org>.
This chapter provides an overview of how to install and learn the software `R`, how to troubleshoot code, and how to perform various data management operations.
Please note that R code is case sensitive—it treats lower case as different from upper case, so you must be very careful to use the correct case when referring to a given object.

@@

1. Download and install `R`: <https://cran.r-project.org>
1. After installing `R`, download and install `RStudio Desktop`—this is a graphical user interface for `R`: <https://posit.co/download/rstudio-desktop>
1. Install tools to allow you to compile `R` packages so you can install packages from source, if necessary (i.e., if package binaries are not available):
    - If on Windows, download and install the version of `RTools` (<https://cran.r-project.org/bin/windows/Rtools/>) that corresponds to your version of `R`
    - If on Mac, download and install `R Compiler Tools for Rcpp on MacOS`: <https://thecoatlessprofessor.com/programming/cpp/r-compiler-tools-for-rcpp-on-macos/>
1. After installing `R` and `RStudio`, set the executables for `R` and `RStudio` to always run with administrator permissions (so they have write access to the main installation directory for installing packages).
    - If on Windows, open File Explorer and find the main executable of `R` (`C:/Program Files/R/[R-VERSION]/bin/R.exe`) and `RStudio` (`C:\Program Files\RStudio\bin\RStudio.exe`).
    Right-click it to open the contextual menu.
    Then, click or tap on "Properties".
    In the Properties window, go to the Compatibility tab.
    At the bottom of the window, check the box next to the "Run this program as an administrator" option, and then click or tap on OK.
1. Open `RStudio` (by right clicking and selecting "Run as Administrator").
1. For [reproducibility purposes](https://devpsylab.github.io/DataAnalysis/R.html#dontSaveWorkspace), prevent `RStudio` from saving your workspaces automatically:
    - With `RStudio` running, choose `Tools → Global Options` from the menus.
    - In the Options dialog, change the value for `Save workspace to .RData on exit` to `Never`.
    - Click `OK`.
1. In `RStudio`, run the following code in the console to install the `R` packages used in this book (note: this will take a while):
   ```{r}
   #| eval: false

>>>>

The book uses the software `R` [@R-base] for statistical analyses (<http://www.r-project.org>).
R is a free software environment; you can download it at no charge here: <https://cran.r-project.org>.
This chapter provides an overview of how to install and learn the software `R`, how to troubleshoot code, and how to perform various data management operations.
Please note that R code is case sensitive—it treats lower case as different from upper case, so you want to be careful to use the correct case when referring to a given object.

@@

1. Download and install `R`: <https://cran.r-project.org>
1. After installing `R`, download and install `RStudio Desktop`—this is a graphical user interface for `R`: <https://posit.co/download/rstudio-desktop>
1. Install tools to allow you to compile `R` packages so you can install packages from source, if necessary (i.e., if package binaries are not available):
    - If on Windows, download and install the version of `RTools` (<https://cran.r-project.org/bin/windows/Rtools/>) that corresponds to your version of `R`
    - If on Mac, download and install `R Compiler Tools for Rcpp on MacOS`: <https://thecoatlessprofessor.com/programming/cpp/r-compiler-tools-for-rcpp-on-macos/>
1. After installing `R` and `RStudio`, set the executables for `R` and `RStudio` to always run with administrator permissions (so they have write access to the main installation directory for installing packages).
    - If on Windows, open File Explorer and find the main executable of `R` (`C:/Program Files/R/[R-VERSION]/bin/R.exe`) and `RStudio` (`C:/Program Files/RStudio/bin/RStudio.exe`).
    Right-click it to open the contextual menu.
    Then, click on "Properties".
    In the Properties window, go to the Compatibility tab.
    At the bottom of the window, check the box next to the "Run this program as an administrator" option, and then click `OK`.
1. Open `RStudio` (by right clicking and selecting "Run as Administrator").
1. For [reproducibility purposes](https://devpsylab.github.io/DataAnalysis/R.html#dontSaveWorkspace), prevent `RStudio` from saving your workspaces automatically:
    - With `RStudio` running, choose `Tools → Global Options` from the menus.
    - In the Options dialog, change the value for `Save workspace to .RData on exit` to `Never`.
    - Click `OK`.
1. In `RStudio`, run the following code in the console to install the `R` packages used in this book (note: this will take a while):
   ```{r}
   #| eval: false
